---
title: Using DualIndexedMatrix objects
author:
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
package: BumpyMatrix
date: "Revised: March 23, 2021"
output:
  BiocStyle::html_document:
    toc_float: yes
vignette: >
  %\VignetteIndexEntry{The DualIndexedMatrix class}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
library(BiocStyle)
set.seed(0)
```

# Motivation

Some matrix representations inherently prefer access on one dimension.
The compressed sparse column format (i.e., `dgCMatrix`) allow for rapid access to random columns but require a series of binary searches to obtain a random row.
Similarly, HDF5 matrices can be chunked so that each row or column is a single chunk, enabling efficient access to either individual rows or columns - but not both at the same time.
This is problematic as typical genomic analysis workflows require iteration across both rows and columns.

For many file formats, the conventional solution is to use a compromise representation that provides good-enough access performance on both dimensions.
However, if disk space/memory is not an issue, another strategy is to simply make two copies of the same dataset for optimal access on each dimension.
One copy is indexed for rapid row access while the other copy is indexed for rapid column access,
and we can dynamically switch between the two depending on the access pattern being used during the analysis.

# The `DualIndexedMatrix` class

The `DualIndexedMatrix` provides a `DelayedArray` subclass that implements this dynamic switching.
Given two matrices with indexing on each dimension, it will automatically use the faster representation for data extraction.
To illustrate, we will create two HDF5-backed copies of the same dataset, one chunked by row and the other chunked by column.

```{r}
library(HDF5Array)
common.mat <- matrix(runif(2000000), ncol=1000)
path <- tempfile(fileext=".hdf5")
byrow <- writeHDF5Array(common.mat, path, name="byrow", chunkdim=c(1, ncol(common.mat)))
bycol <- writeHDF5Array(common.mat, path, name="bycol", chunkdim=c(nrow(common.mat), 1))
```

Construction of the `DualIndexedMatrix` is straightforward.
We set `row=` to a matrix (or the equivalent `DelayedArray` seed) that supports fast row access,
and we set `column=` to another matrix that supports fast column access.

```{r}
library(DualIndexedMatrix)
mat <- DualIndexedMatrix(row=HDF5ArraySeed(path, "byrow"), 
    column=HDF5ArraySeed(path, "bycol"))
mat
```

And that's it!

# Demonstrating performance

To compare, let's iterate over rows and compute the row sums.
We'll use the `r Biocpkg("beachmat")` package for convenience and we'll lower the block sizes to get more iterations.
(Otherwise, the matrix would be too small to see any difference in performance.)

```{r}
library(beachmat)
setAutoBlockSize(1e5)

system.time(rowBlockApply(byrow, FUN=rowSums))
system.time(rowBlockApply(bycol, FUN=rowSums))
system.time(rowBlockApply(mat, FUN=rowSums))
```

Similarly, we iterate over columns and compute the column sums.

```{r}
system.time(colBlockApply(byrow, FUN=colSums))
system.time(colBlockApply(bycol, FUN=colSums))
system.time(colBlockApply(mat, FUN=colSums))
```

We can see that the `DualIndexedMatrix` performs well in both dimensions while the others can only provide fast access in a single dimension.
This is because the former can just switch between the latter two depending on the access pattern.

Of course, this particular example is somewhat pathological as dense matrices are best stored in HDF5 with rectangular chunks.
This gives decent performance in our example without needing to duplicate the dataset.

```{r}
rect <- writeHDF5Array(common.mat, path, name="square", chunkdim=c(100, 100))
system.time(rowBlockApply(rect, FUN=rowSums))
system.time(colBlockApply(rect, FUN=colSums))
```

Even so, the `DualIndexedMatrix` still exhibits a modest improvement. 
This advantage can be expected to increase for non-consecutive row/column access where rectangular chunks are less efficient.
The dual approach is also highly effective for formats that are inherently biased towards one dimension, e.g., the compressed sparse representation used in 10X Genomics output and H5AD files.

# Internal transposition

Advanced users can instruct the `DualIndexedMatrix` to dynamically transpose the seed contents during data extraction.
This is mostly intended for cases where the implementation of the format clearly prefers access along one dimension.
So, rather than trying to create a new implementation for the other dimension, we can transpose the data, 
store it in the format of interest, and then un-transpose the data during extraction.

The simplest example is that of the `dgCMatrix` class, which is the most well-supported sparse representation in the `r CRANpkg("Matrix")` package.
(A `dgRMatrix` class technically exists but is rarely used.)
If we want fast row access, the easiest solution is to transpose our input matrix so that obtaining a row involves quickly extracting a column from the transposed matrix.
We can indicate that this is the case by setting `row.transposed=TRUE`, as shown below.

```{r}
library(Matrix)
by.column <- rsparsematrix(100, 200, 0.1)
by.row <- t(by.column) 
mat <- DualIndexedMatrix(by.row, by.column, row.transposed=TRUE) 
mat
```

# Closing remarks

The general philosophy here is to spend memory/disk space to improve access speed.
Data providers pay an upfront cost to index the dataset in both dimensions, which users can exploit by creating a `DualIndexedMatrix` for their analyses.
Indeed, even if the data provider _is_ the user, faster access throughout the lifetime of an analysis project will usually outweigh any indexing costs.
This is most relevant for large datasets that require file-backed representations and time-consuming disk I/O, where the benefits of optimized access are more apparent.

On that note: file formats like HDF5 are particularly convenient for this strategy, as both copies of the dataset can be stored in the same file.
This simplifies the management of the dataset, e.g., downloading, caching and synchronization of the dual indexed copies.
However, any format or file organisation can be used as long as the data is the same in both copies.
In fact, one can even use combinations of file-backed or in-memory formats in the `DualIndexedMatrix`, depending on what is available.

# Session information {-}

```{r}
sessionInfo()
```
